---
title: "Readme"
author: "Julie I. Santos"
date: "Wednesday, March 18, 2015"
output: html_document
---

From the instructions: The code should have a file run_analysis.R in the main directory that can be run as long as the Samsung data is in your working directory.  I chose to assume that the files were in the correct directory.

First, I read in the test data from the three files and bind them together with a factor variable called groupID that will help subsequent analysis determine whether the data came from the test or training group.

```{r}
library(dplyr)
sub <- read.table("./test/subject_test.txt")
x <- read.table("./test/X_test.txt")
y <- read.table("./test/Y_test.txt")
test <- cbind(sub,x,y, groupId = as.factor("TEST"))
```

Similarly with the training data:

```{r}
sub <- read.table("./train/subject_train.txt")
x <- read.table("./train/X_train.txt")
y <- read.table("./train/Y_train.txt")
train <- cbind(sub,x,y, groupId = as.factor("TRAIN"))
```

I then combined the data with rbind
```{r}
data <- rbind(test,train)
```

The *features.txt* file contains most of the column names. However, it reads in with an index, so I grab the second column and combine it with the strings "subject", "activity", and "groupID" to get the column names.  Since the names in *features.txt* have parens, they won't make valid column names, so I use the function make.names().

```{r}
n <- read.table("features.txt")
n <- as.character(n[,2])
n <- c("subject",n, "activity","groupID")
n <-  make.names(names=n, unique=TRUE, allow_ = TRUE)
colnames(data) <- n
```

I made a new column to translate the numeric *activity* variable to a factor with six levels.

```{r}
data$activities <- factor(data$activity, labels=c("WALKING", "WALKING_UPSTAIRS","WALKING_DOWNSTAIRS","SITTING","STANDING","LAYING"))
```

The function *make.names()* took the parentheses out of the column names and replaced them with periods.  From the 
You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
